---
title: "ComBat Family Overview"
author: "Andrew Chen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{ComBat Family Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{neuroCombat, mgcv, gamlss, longCombat, lme4}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The ComBat Family extends the original ComBat methodology to enable flexible
covariate modeling, leveraging efficient R implementations of regression models.
A method that belongs in the ComBat Family satisfies the following conditions:

1. Modeling of covariate effects in location and/or scale
2. Batch effects in location and scale of measurements
3. Empirical Bayes step for borrowing information across features

Denote data $y_{ijv}$ and covariates $\boldsymbol{x}_{ij}$ for sites $i = 1,2,\ldots,M$, subjects within site $j = 1,2,\ldots,n_i$, and features $v = 1,2,\ldots,p$. The ComBat Family assumes 

$$y_{ijv} \sim D(\mu_{ijv} + \gamma_{iv},  \delta_{iv}\sigma_{ijv})$$
where $D$ is a specified distribution (typically normal), $\gamma_{iv}$ is the additive location batch effect and $\delta_{iv}$ is the multiplicative scale batch effect. The mean and variance of $D$ are modeled via

$$g_1(\mu_{ijv}) = f_v(\boldsymbol{x}_{ij}) \qquad g_2(\sigma_{ijv}) = h_v(\boldsymbol{x}_{ij})$$

where $g_1, g_2$ are link functions and $f_v, h_v$ are specified functions.

## ComBat
ComBat Family defaults to the original linear model used in ComBat, using 
`formula` to specify covariates included in the model.

```{r}
suppressPackageStartupMessages({
  library(neuroCombat)
  library(ComBatFamily)
})

# generate toy dataset
set.seed(8888)
n <- 20
p <- 5
bat <- as.factor(c(rep("a", n/2), rep("b", n/2)))
q <- 2
covar <- matrix(rnorm(n*q), n, q)
colnames(covar) <- paste0("x", 1:q)
data <- data.frame(matrix(rnorm(n*p), n, p))

cf <- comfam(data, bat, covar, lm, formula = y ~ x1 + x2)
c <- neuroCombat(t(data), bat, covar, verbose = FALSE)

max(cf$dat.combat - t(c$dat.combat))
```

```{r}
# no covariates specified
cf <- comfam(data, bat)
c <- neuroCombat(t(data), bat, verbose = FALSE)

max(cf$dat.combat - t(c$dat.combat))
```

## ComBat-GAM
ComBat-GAM (Pomponio et al., 2020) was previously only available in Python. 
ComBat Family enables fitting of a general additive model (GAM) using notation
consistent with the `gam` function.

```{r}
suppressPackageStartupMessages(
  library(mgcv)
)
cg <- comfam(data, bat, covar, gam, formula = y ~ s(x1) + x2)
```


## ComBat with location-scale models (ComBat-LS)
ComBat Family supports modeling of covariate effects on the scale of
measurements via GAMLSS, which is still experimental. Testing is ongoing, but 
ComBat-LS should be considered in settings with potential covariate effects
on scale (e.g. age affecting the variance of imaging features).

```{r}
suppressPackageStartupMessages(
  library(gamlss)
)

cgl <- comfam(data, bat, covar, gamlss, y ~ x1 + x2,
              sigma.formula = ~ x1 + x2, 
              control = gamlss.control(trace = FALSE))
```

## Longitudinal ComBat
The ComBat family includes Longitudinal ComBat (Beer et al. 2020), previously implemented [here](https://github.com/jcbeer/longCombat). Currently, `comfam` uses the
mean squared residual (MSR) estimator for error variance, but the REML estimator will be implemented at a later date.

```{r}
# devtools::install_github("jcbeer/longCombat")
suppressPackageStartupMessages({
  library(longCombat)
  library(lme4)
})

# generate toy dataset
n <- 20
p <- 5
r <- 2 # repeats
bat <- as.factor(rep(c(rep("a", n/2), rep("b", n/2)), r))
q <- 2
covar <- matrix(rnorm(n*r*q), n*r, q)
colnames(covar) <- paste0("x", 1:q)
covar <- cbind(covar, ID = rep(1:n, r))
data <- data.frame(matrix(rnorm(n*r*p), n*r, p))

cf <- comfam(data, bat, covar, lmer, y ~ x1 + x2 + (1 | ID))
c <- longCombat("ID", "x1", "bat", 1:p, "x1 + x2", "(1 | ID)",
                data = cbind(data, covar, bat), method = "MSR", verbose = FALSE)

max(cf$dat.combat - c$data_combat[,-(1:3)])
```

## References
Beer, J. C., Tustison, N. J., Cook, P. A., Davatzikos, C., Sheline, Y. I., Shinohara, R. T., & Linn, K. A. (2020). Longitudinal ComBat: A method for harmonizing longitudinal multi-scanner imaging data. *NeuroImage*, 220, 117129. https://doi.org/10.1016/j.neuroimage.2020.117129

Johnson, W. E., Li, C., & Rabinovic, A. (2007). Adjusting batch effects in microarray expression data using empirical Bayes methods. *Biostatistics*, 8(1), 118–127. https://doi.org/10.1093/biostatistics/kxj037

Pomponio, R., Erus, G., Habes, M., Doshi, J., Srinivasan, D., Mamourian, E., Bashyam, V., Nasrallah, I. M., Satterthwaite, T. D., Fan, Y., Launer, L. J., Masters, C. L., Maruff, P., Zhuo, C., Völzke, H., Johnson, S. C., Fripp, J., Koutsouleris, N., Wolf, D. H., … Davatzikos, C. (2020). Harmonization of large MRI datasets for the analysis of brain imaging patterns throughout the lifespan. *NeuroImage*, 208, 116450. https://doi.org/10.1016/j.neuroimage.2019.116450


