---
title: "ComBat Family Overview"
author: "Andrew Chen"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{ComBat Family Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The ComBat Family extends the original ComBat methodology to enable flexible
covariate modeling, leveraging efficient R implementations of regression models.
A method that belongs in the ComBat Family satisfies the following conditions:

1. Modeling of covariate effects in location and scale
2. Identity link function, data on the support $(-\infty, \infty)$
3. Batch effects in location and scale of measurements
4. Empirical Bayes step for borrowing information across features

## Original ComBat
ComBat Family readily supports the original linear model used in ComBat via
`model = lm` and using `formula` to specify covariates included in the model.

```{r}
suppressPackageStartupMessages({
  library(neuroCombat)
  library(ComBatFamily)
})

# generate toy dataset
set.seed(8888)
n <- 20
p <- 100
bat <- as.factor(c(rep("a", n/2), rep("b", n/2)))
q <- 2
covar <- matrix(rnorm(n*q), n, q)
colnames(covar) <- paste0("x", 1:q)
data <- data.frame(matrix(rnorm(n*p), n, p))

cf <- comfam(data, covar, bat, lm, formula = y ~ x1 + x2)
c <- neuroCombat(t(data), bat, covar, verbose = FALSE)

max(cf$dat.combat - t(c$dat.combat))
```

## ComBat-GAM
ComBat-GAM (Pomponio et al., 2020) was previously only available in Python. 
ComBat Family enables fitting of a general additive model (GAM) using notation
consistent with the `gam` function.

```{r}
suppressPackageStartupMessages(
  library(mgcv)
)
cg <- comfam(data, covar, bat, gam, formula = y ~ s(x1) + x2)
```

## Longitudinal ComBat
The ComBat family includes Longitudinal ComBat (Beer et al. 2020), previously implemented [here](https://github.com/jcbeer/longCombat). Currently, `comfam` uses the
mean squared residual (MSR) estimator for error variance, but the REML estimator will be implemented at a later date.

```{r}
# devtools::install_github("jcbeer/longCombat")
suppressPackageStartupMessages({
  library(longCombat)
  library(lme4)
})

# generate toy dataset
n <- 20
p <- 10
r <- 2 # repeats
bat <- as.factor(rep(c(rep("a", n/2), rep("b", n/2)), r))
q <- 2
covar <- matrix(rnorm(n*r*q), n*r, q)
colnames(covar) <- paste0("x", 1:q)
covar <- cbind(covar, ID = rep(1:n, r))
data <- data.frame(matrix(rnorm(n*r*p), n*r, p))

cf <- comfam(data, covar, bat, lmer, y ~ x1 + x2 + (1 | ID))
c <- longCombat("ID", "x1", "bat", 1:p, "x1 + x2", "(1 | ID)",
                data = cbind(data, covar, bat), method = "MSR", verbose = FALSE)

max(cf$dat.combat - c$data_combat[,-(1:3)])
```

## ComBat-GAMLSS
ComBat Family supports modeling of covariate effects on the scale of
measurements via GAMLSS, which is still experimental. Further testing is 
warranted, but ComBat-GAMLSS should be considered in settings with known 
covariate effects on scale.

```{r}
suppressPackageStartupMessages(
  library(gamlss)
)

n <- 20
p <- 10
bat <- as.factor(c(rep("a", n/2), rep("b", n/2)))
q <- 2
covar <- matrix(rnorm(n*q), n, q)
colnames(covar) <- paste0("x", 1:q)
data <- data.frame(matrix(rnorm(n*p), n, p))

cgl <- comfam(data, covar, bat, gamlss, y ~ x1 + x2,
              sigma.formula = ~ x1 + x2, 
              control = gamlss.control(trace = FALSE))
```

## References
Beer, J. C., Tustison, N. J., Cook, P. A., Davatzikos, C., Sheline, Y. I., Shinohara, R. T., & Linn, K. A. (2020). Longitudinal ComBat: A method for harmonizing longitudinal multi-scanner imaging data. *NeuroImage*, 220, 117129. https://doi.org/10.1016/j.neuroimage.2020.117129

Johnson, W. E., Li, C., & Rabinovic, A. (2007). Adjusting batch effects in microarray expression data using empirical Bayes methods. *Biostatistics*, 8(1), 118–127. https://doi.org/10.1093/biostatistics/kxj037

Pomponio, R., Erus, G., Habes, M., Doshi, J., Srinivasan, D., Mamourian, E., Bashyam, V., Nasrallah, I. M., Satterthwaite, T. D., Fan, Y., Launer, L. J., Masters, C. L., Maruff, P., Zhuo, C., Völzke, H., Johnson, S. C., Fripp, J., Koutsouleris, N., Wolf, D. H., … Davatzikos, C. (2020). Harmonization of large MRI datasets for the analysis of brain imaging patterns throughout the lifespan. *NeuroImage*, 208, 116450. https://doi.org/10.1016/j.neuroimage.2019.116450


