---
title: "ComBat Family Overview"
author: "Andrew Chen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{ComBat Family Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The CovBat Family extends the original CovBat methodology to enable flexible
covariate modeling, leveraging efficient R implementations of regression models.
A method that belongs in the CovBat Family satisfies the following conditions:

1. Modeling of covariate effects in location and scale
2. Real-valued measurements and identity link function
3. Batch effects in mean and covariance of measurements
4. Empirical Bayes step for borrowing information across features
5. Adjustment of multivariate covariate and batch effects through principal 
components (PCs) of ComBat residuals

## CovBat
CovBat Family defaults to the original linear model and parameters used in 
[CovBat](https://github.com/andy1764/CovBat_Harmonization)

```{r}
suppressPackageStartupMessages({
  library(CovBat)
  library(ComBatFamily)
})

# generate toy dataset
set.seed(8888)
n <- 20
p <- 5
bat <- as.factor(c(rep("a", n/2), rep("b", n/2)))
q <- 2
covar <- matrix(rnorm(n*q), n, q)
colnames(covar) <- paste0("x", 1:q)
data <- data.frame(matrix(rnorm(n*p), n, p))

cf <- covfam(data, bat, covar, lm, formula = y ~ x1 + x2)
c <- covbat(t(data), bat, covar)

max(cf$dat.covbat - t(c$dat.covbat))
```

```{r}
# no covariates specified
cf <- covfam(data, bat)
c <- covbat(t(data), bat)

max(cf$dat.covbat - t(c$dat.covbat))
```

## CovBat-GAM
Modeling covariates via a general additive model (GAM) is part of the CovBat
family and can be easily implemented.

```{r}
suppressPackageStartupMessages(
  library(mgcv)
)
cg <- covfam(data, bat, covar, gam, formula = y ~ s(x1) + x2)
```

## CovBat with Modeling of Scores
CovBat Family extends the original CovBat method by enabling modeling of
covariate effects in the principal component (PC) scores. The model is estimated
separately from the model used in the standardization step.

```{r}
# no covariates specified
cf <- covfam(data, bat, covar, formula = y ~ x1 + x2, score.model = lm,
             score.args = list(formula = y ~ x1 + x2))
```

## References
Chen, A. A., Beer, J. C., Tustison, N. J., Cook, P. A., Shinohara, R. T., Shou, H., & Initiative, T. A. D. N. (2022). Mitigating site effects in covariance for machine learning in neuroimaging data. *Human Brain Mapping*, 43(4), 1179–1195. https://doi.org/10.1002/hbm.25688

